### 2023-12-17 tesis_Ashish/Chapter3/chapter3.tex
update
thesis version 11 Jul
Sehrawat A thesis 17 Dec 2023

\chapter{PCC luminometer calibration}  %Title of the First Chapter

@@

%One of the widely used method for the calibration of luminosity measurement devices is the van der Meer (VdM) scan method.
A prevalent technique for calibrating luminosity measurement devices is the van der Meer (VdM) scan method. This method is named after Simon van der Meer, a Dutch physicist and Nobel laureate, who developed the technique in the 1960s. These scans are performed to relate the collision rate of a given luminometer to the absolute luminosity determined from the measured beam parameters. The proportionality constant is called the visible cross section. It involves scanning the beams across each other in the horizontal and vertical planes to measure the beam overlap integral, which is proportional to the luminosity. The first step in the VdM calibration is to scan the colliding beams across each other. This is done by displacing the beams in the horizontal (x) and vertical (y) planes independently while keeping the other direction fixed as shown in left part of Fig. \ref{fig:vdm_scan_method}. The scans are performed at several different positions along the collision point, creating a series of data points for each plane. The data points collected during the scans are then fitted with Gaussian function or other suitable function like double gaussian, Poly2G, Poly4G, etc as shown in right part of Fig. \ref{fig:vdm_scan_method}  to determine the beam overlap integrals in the x and y directions. The product of the integrals in the two planes yields the total beam overlap integral. %The event rate is proportional to the luminosity and is given by the product of the cross-section ($\sigma$) and luminosity. 
The beam overlap integral is the integral of the product of the beam overlap width in the horizontal and vertical planes over the entire interaction region \cite{vanderMeer1968}. It is an essential parameter in determining the absolute luminosity.

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{ashish_thesis/vdm_method.png}
    \caption[Schematic of van der Meer scan]{Left: Scanning the beams across each other (±3$\sigma$) in the transverse plane, horizontal (x-scan) and vertical (y-scan) directions. Right: Rate is measured as a function of beam separation}
    \label{fig:vdm_scan_method}
\end{figure}

@@

In the van der Meer (vdM) scans,  rate is plotted as a function of beam overlap widths along x and y directions that is used to determine the visible cross section $\sigma_{vis}$ of the pixel detector which is a calibration constant in the relation between rate of a quantity (clusters, coincidences) read out by the detector and the instantaneous luminosity.

@@

\begin{equation}
L_{b} = N_1 N_2 f \int^{+\infty}_{-\infty} \rho_1(x,y) \rho_2(x+\Delta x, y+\Delta y) dx dy 
\end{equation}

where $N_1, N_2$ are the number of protons in the colliding bunches, f is the LHC orbit frequency whose value is 11246Hz,  $\rho_1(x,y)$ and $\rho_2(x+\Delta x,y+\Delta y)$ are the two dimensional particle density distribution functions for each colliding bunch separated by $\Delta x$ and $\Delta y$ in x and y directions  \cite{CMS-PAS-LUM-17-004}.

@@

\begin{equation}
L_{b} = N_1 N_2 f \int^{+\infty}_{-\infty} \rho_{1x}(x) \rho_{2x}(x+\Delta x)  dx   \int^{+\infty}_{-\infty} \rho_{1y}(y) \rho_{2y}(y+\Delta y)  dy 
\end{equation}

where $\rho_1(x,y) = \rho_{1x}(x) \rho_{1y} (y)$ and $\rho_2(x,y) = \rho_{2x}(x+\Delta x) \rho_{2y} (y+\Delta y)$

The vdM scan determines the bunch  overlap integrals $\int \rho_{x1} (x) \rho_{x2} (x+\Delta x) dx$ and  $\int \rho_{y1}(y) \rho_{y2} (y+\Delta y) dy$ by measuring pixel clusters rate as a function of bunch separation $\Delta x$ and $\Delta y$ along x and y directions as shown in Fig. 14.

\begin{equation}
\int \rho_{x1} (x) \rho_{x2} (x+\Delta x) dx = \frac{R_x(0)}{\int R_x(\Delta x)dx} 
\end{equation}

\begin{equation}
\int \rho_{y1} (y) \rho_{y2} (y+\Delta y) dy = \frac{R_y(0)}{\int R_y(\Delta y)dy} 
\end{equation}

where $R_x(0)$ is the measured cluster rate along x direction when the bunch separation is zero and $R_x(\Delta x)$ is the measured cluster rate when bunches are separated by distance $\Delta x$ and $R_y(0)$ is the measured cluster rate along y direction when the bunch separation is zero and $R_y(\Delta y)$ is the measured cluster rate when bunches are separated by distance $\Delta y$.

Bunch overlap widths $\Sigma_x$ and $\Sigma_y$ along x and y directions are defined as

\begin{equation}
\Sigma_x = \frac{1}{\sqrt{2 \pi}} \frac{\int R_x(\Delta x)dx}{R_x(0)} 
\end{equation}

\begin{equation}
\Sigma_y =  \frac{1}{\sqrt{2 \pi}} \frac{\int R_y(\Delta y)dy}{R_y(0)} 
\end{equation}

@@

where R(0) is the peak cluster rate measured by the pixel detector after applying appropriate stability module selection.

@@

The calibration of luminosity measurement by the CMS experiment in the 2018 proton-proton data taking at √s = 13 TeV was performed during

LHC fill 6868 on June 30 and July 1, 2018 at √s = 13 TeV

Zero-bias triggers on 5 bunch pairs (BCIDs 265, 865, 1780, 2192, and 3380)

recorded events rate is 27.7 kHz.

vdM scans are typically conducted at least once a year to ensure the luminometer is correctly calibrated. Results of these scans are used to normalize the data collected by the experiments during physics run.

@@

\section{Background Estimation}

@@

\section{vdM fit model}

The poly2G model, or the "Gaussian function with a polynomial term" model, is a fitting function that combines a second-order polynomial function and a Gaussian function. It is a versatile and flexible model that can be used to fit various types of data, including the PCC rate plot in the context of the Van der Meer (VdM) calibration process. In a PCC rate plot, the event rate is plotted as a function of the beam separation. The poly2G model can be employed to fit the PCC rate plot to extract the beam parameters, such as the beam overlap widths and the center positions, which are essential for the luminosity measurement \cite{Cekmecelioglu:2775639}. The general form of the poly2G model is

@@

\begin{equation}
Poly2G = Peak  \left [1+r2 \left(\frac{x-\mu} {\frac{\sigma}{1+r2}}\right)^2 \right] exp \left[-\frac{1}{2}\left(\frac{x-\mu}{\frac{\sigma}{1+r2}}\right)^2\right]
\end{equation}

where Peak is the maximum PCC rate, $\mu$ is the mean, $\sigma$ and r2 are the fit parameters.

>>>>

\chapter{Luminometer calibration}  %Title of the First Chapter

@@

%One of the widely used method for the calibration of luminosity measurement devices is the van der Meer (VdM) scan method.
A prevalent technique for calibrating luminosity measurement devices is the van der Meer (vdM) scan method. This method is named after Simon van der Meer, a Dutch physicist and Nobel laureate, who developed the technique in the 1960s. These scans are performed to relate the detection rate of a given luminometer to the absolute luminosity determined from the measured beam parameters. The proportionality constant is called the visible cross section. It involves scanning the beams across each other in the horizontal and vertical planes to measure the beam overlap integral, which is proportional to the luminosity. The first step in the vdM calibration is to scan the colliding beams across each other. This is done by displacing the beams in the horizontal (x) and vertical (y) planes independently while keeping the other direction fixed as shown in left part of Fig. \ref{fig:vdm_scan_method}. The scans are performed at several different positions along the collision point, creating a series of data points for each plane. The data points collected during the scans are then fitted with  suitable function like single, double gaussian as shown in right part of Fig. \ref{fig:vdm_scan_method}  to determine the beam overlap integrals in the x and y directions. %The product of the integrals in the two planes yields the total beam overlap integral. The event rate is proportional to the luminosity and is given by the product of the cross-section ($\sigma$) and luminosity. 
The total beam overlap is the product of beam overlap width in the horizontal and vertical planes over the entire interaction region \cite{vanderMeer1968}. It is an essential parameter in determining the absolute luminosity to calculate visible cross section.

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{ashish_thesis/vdm_method.png}
    \caption[vdM Scan Schematic]{Left: Scanning the beams across each other (±3$\sigma$) in the transverse plane, horizontal (x-scan) and vertical (y-scan) directions. Right: Rate is measured as a function of beam separation \cite{karacheban2017luminosity}.}
    \label{fig:vdm_scan_method}
\end{figure}

@@

In the van der Meer (vdM) scans,  rate is plotted as a function of beam separation along x and y directions that is used to determine the visible cross section $\sigma_{vis}$.
%of the pixel detector which is a calibration constant in the relation between rate of a quantity (clusters, coincidences) read out by the detector and the instantaneous luminosity.

@@

\begin{equation}
L_{b} = N_1 N_2 f \int^{+\infty}_{-\infty} \rho_1(x,y) \rho_2(x, y) dx dy 
\end{equation}

where $N_1, N_2$ are the number of protons in the colliding bunches, f is the LHC orbit frequency whose value is 11246 Hz,  $\rho_1(x,y)$ and $\rho_2(x,y)$ are the two dimensional particle density distribution functions for each colliding bunch in x and y directions  \cite{CMS-PAS-LUM-17-004}.

@@

\begin{equation}
L_{b} = N_1 N_2 f \int^{+\infty}_{-\infty} \rho_{1x}(x) \rho_{2x}(x)  dx   \int^{+\infty}_{-\infty} \rho_{1y}(y) \rho_{2y}(y)  dy 
\end{equation}

where $\rho_1(x,y) = \rho_{1x}(x) \rho_{1y} (y)$ and $\rho_2(x,y) = \rho_{2x}(x) \rho_{2y} (y)$

The vdM scan determines the bunch  overlap integrals $\int \rho_{1x} (x) \rho_{2x} (x) dx$ and  $\int \rho_{1y}(y) \rho_{2y} (y) dy$ by measuring pixel clusters rate as a function of beam separation $\Delta x$ and $\Delta y$ along x and y directions.

\begin{equation}
\int \rho_{1x} (x) \rho_{2x} (x) dx = \frac{R_x(0)}{\int R_x(\Delta x)d \Delta x} 
\end{equation}

\begin{equation}
\int \rho_{1y} (y) \rho_{2y} (y) dy = \frac{R_y(0)}{\int R_y(\Delta y)d \Delta y} 
\end{equation}

where $R_x(0)$ is the measured cluster rate along x direction when the beam separation is zero and $R_x(\Delta x)$ is the measured cluster rate when beams are separated by distance $\Delta x$ and $R_y(0)$ is the measured cluster rate along y direction when the beam separation is zero and $R_y(\Delta y)$ is the measured cluster rate when beams are separated by distance $\Delta y$.

We define the quantities $\Sigma_x$ and $\Sigma_y$:

%Bunch overlap widths $\Sigma_x$ and $\Sigma_y$ along x and y directions are defined as \cite{White:1308187}

\begin{equation}
\Sigma_x = \frac{1}{\sqrt{2 \pi}} \frac{\int R_x(\Delta x)d \Delta x}{R_x(0)} 
\end{equation}

\begin{equation}
\Sigma_y =  \frac{1}{\sqrt{2 \pi}} \frac{\int R_y(\Delta y)d \Delta y}{R_y(0)} 
\end{equation}

@@

where R(0) is the peak cluster rate. % measured by the pixel detector after applying appropriate stability module selection.

@@

\begin{comment}
The calibration of luminosity measurement by the CMS experiment in the 2018 proton-proton data taking at √s = 13 TeV was performed during

\begin{itemize}
\item LHC fill 6868 on June 30 and July 1, 2018 at √s = 13 TeV

\item Zero-bias triggers on 5 bunch pairs (BCIDs 265, 865, 1780, 2192, and 3380)

\item recorded events rate is 27.7 kHz. 
\end{itemize}

vdM scans are typically conducted at least once a year to ensure the luminometer is correctly calibrated. Results of these scans that is the visible cross section is used to normalize the data collected by the experiments during physics run.

@@

%\begin{figure}[!htp]
%\centering
%\includegraphics[width=1\textwidth]{ashish_thesis/new_framework_vdm.png}
%\caption{%
%   klm
%}
%\label{fig:vdm_new_f}
%\end{figure}

\section{Background estimation}

@@

\section{Fit model}

The poly2G model, or the "Gaussian function with a polynomial term" model, is a fitting function that combines a second-order polynomial function and a Gaussian function. It is a versatile and flexible model that can be used to fit various types of data, including the PCC data in the context of the vdM calibration. %In a PCC rate plot, the rate is plotted as a function of the beam separation.
The Poly2G model can be employed to fit the PCC rates %to display data and fit model in the plot
and extract the beam parameters, beam overlap integrals and peak rates  needed for the determination of the calibration constant  %such as the beam overlap widths and the beam center positions, which are essential for the luminosity measurement
\cite{Cekmecelioglu:2775639}. The Poly2G function has additional degrees of freedom compared to a single Gaussian but fewer than a DG (Double Gaussian), allowing for better convergence. The general form of the Poly2G model is

@@

\begin{equation}
F(x) = Peak  \left [1+r_2 \left(\frac{\Delta x-\mu} {\frac{\sigma}{1+r_2}}\right)^2 \right] exp \left[-\frac{1}{2}\left(\frac{\Delta x-\mu}{\frac{\sigma}{1+r_2}}\right)^2\right]
\end{equation}

where Peak is the maximum PCC rate, $\mu$ is the mean, $\sigma$ and $r_2$ are the fit parameters.

To determine the beam overlap $\Sigma$ in this model, we calculate the area under the Poly2G curve between $-\infty$ and  +$\infty$ and divide by its peak value. We integrate the poly2G function
%\[
%F(x) = \text{Peak} \left[ 1 + r_2 \left( \frac{x - \mu}{\frac{\sigma}{1 + r_2}} \right)^2 \right] \exp \left[ -\frac{1}{2} \left( \frac{x - \mu}{\frac{\sigma}{1 + r_2}} \right)^2 \right]
%\]
from \(-\infty\) to \(+\infty\) using the substitution method.

\begin{equation}
z = \frac{x - \mu}{\frac{\sigma}{1 + r_2}} ,   dx = \frac{\sigma}{1 + r_2} dz 
\end{equation}

With this substitution, the function becomes:

\begin{equation}
F(z) = \text{Peak} \left[ 1 + r_2 z^2 \right] \exp\left(-\frac{1}{2} z^2\right)
\end{equation}

The integral is then:

\begin{equation}
I = \text{Peak} \int_{-\infty}^{\infty} \left[ 1 + r_2 z^2 \right] \exp\left(-\frac{1}{2} z^2\right) \frac{\sigma}{1 + r_2} dz
\end{equation}

%Breaking this into two parts:

%\begin{equation}  
%I_1 & = \text{Peak} \int_{-\infty}^{\infty} \exp\left(-\frac{1}{2} z^2\right) \frac{\sigma}{1 + r_2} dz = \text{Peak} \sqrt{2\pi} \frac{\sigma}{1 + r_2} \\
%\end{equation}

%\begin{equation}
%I_2 & = \text{Peak} r_2 \int_{-\infty}^{\infty} z^2 \exp\left(-\frac{1}{2} z^2\right) \frac{\sigma}{1 + r_2} dz = \text{Peak} r_2 \sqrt{2\pi} \frac{\sigma}{1 + r_2}
%\end{equation}

Thus:
\[ I = \frac{\text{Peak} \sqrt{2\pi} \sigma (1 + r_2) }{1+r_2} = \text{Peak} \sqrt{2\pi} \sigma \]

Dividing by the \(\text{Peak}\) value:
\[ \frac{I}{\text{Peak}} = \sqrt{2\pi} \sigma  \]

Beam overlap width for Poly2G fit function is
\begin{equation}
 \Sigma = \frac{ \sqrt{2\pi} \sigma}{\sqrt{2 \pi}} = \sigma 
\end{equation}
